---
title: "Analysis of Means Assignment"
author: "Candice Weber"
format: html
abstract: "Utilizing analysis of variance to understand the effects of gender and diet on weight loss."

---

Here is some new data to play with a bit to explore one-way and two-way analysis of variance models.  These data represent weight data associated with three different diet types.  There are 76 observations of individuals with the following characteristics:

- Gender (0/1)
- Age (earth years. 🤓)
- Height (cm)
- Pre.weight & weight6weeks (kg)
- Diet (1,2,3)

The data are located as a CSV file named `DietData.csv` linked from the Canvas page.  Load the data in and format things so they look correctly to you.

```{r message = FALSE}
library(tidyverse)
library(knitr)

data <- read_csv("DietData.csv")

data %>% 
  mutate(weight.delta = pre.weight - weight6weeks) %>% 
  mutate(Diet = case_when(
        Diet == 1 ~ "Diet 1",
        Diet == 2 ~ "Diet 2",
        Diet == 3 ~ "Diet 3")) -> data
data$gender <- as.factor(data$gender)
```
## Exploratory Data Anlsysis

One of the first things to do is to look at the data and see if there are any obvious things.  Go ahead and explore these data visually.  What do you see?

```{r}

# Response variable: weight loss (difference between pre-weight and weight after 6 weeks in kg)
# Predictor variable: gender (gender 0 or 1)
# Predictor variable: diet (Diet 1, Diet 2, or Diet 3)

ggplot(data, aes(x=gender, y=weight.delta, fill=gender)) + 
  geom_boxplot(notch = TRUE) + 
  labs(x = "Gender",
       y = "Weight Loss",
       title = "Variance is not entirely equal") +
  scale_fill_manual(values = c("purple", "green"))+
  guides(fill = "none")+
  theme_bw()
  
ggplot(data, aes(x=Diet, y=weight.delta, fill = Diet)) +
  geom_boxplot(notch = TRUE) +
  labs(x = "Diet",
       y = "Weight Loss",
       title = "Variance is not entirely equal") +
  scale_fill_manual(values = c("lightblue", "blue", "navy")) +
  guides(fill = "none") +
  theme_bw()

data %>% ggplot(aes(x = weight.delta))+
  geom_histogram(bins = 15, fill = "orange", color = "black")+
  labs(x = "Weight Loss (kg)",
       y = "Frequency",
       title = "Response variable is not super normal")+
  theme_bw()


```



## Estimating Mean Values

Make a table of Weight Loss by gender and diet.

```{r}
data %>%
  group_by(gender, Diet) %>%
  summarize(mean_value = mean(weight.delta, na.rm = TRUE), .groups = "drop") %>%
  spread(key = Diet, value = mean_value, fill = NA) -> summary_table


summary_table %>%
  kable( caption = "Mean Weight Loss (kg) by Gender and Diet",
         digits = 2)

```


## 1-Way Analysis of Variance

*Diet Issues:* 


The underlying linear model.


$$
y_{ij} = \mu + \tau_{Diet, i} + \epsilon_j
$$


Test the null hypothesis, $H_O:$ *There is no effect of diet on weight loss* (e.g., $\tau_{Diet-1} = \tau_{Diet-2} = \tau_{Diet-3} = 0.0$).  Is there evidence for one diet producing more weight loss than the others?  Create an `aov` analysis and assign it to the variable `fit.diet` and examine its contents.

```{r}
fit.diet <- aov(weight.delta ~ Diet, data = data)
anova(fit.diet)
```

Are they all significantly different?  Try the `TukeyHSD()` Interpret the results.

```{r}
diet.tukey <- TukeyHSD(fit.diet)
diet.tukey

diet.sum <- data %>% 
  group_by(Diet) %>% 
  summarise(MEAN = mean(weight.delta),
            N = length(weight.delta),
            SD = sd(weight.delta),
            CI= 1.96 * sd(weight.delta) / sqrt(length(weight.delta)))
diet.sum
```
 
 
 
<span style="color: red;">Based on this ANOVA, I reject the null hypothesis (F = 5.383, DF = 2, p = 0.007) and conclude that the average weight loss of at least one of the diets differs from the overall average weight loss.  The mean weight loss with diet 3 (5.148 +/- 0.904) was greater than the mean weight loss with diet 1 (3.300 +/- 0.896) or diet 2 (3.268 +/- 0.966).  The post-hoc Tukey test did find significant differences between the mean weight loss of diet 3 and diet 2 (p = 0.152) and diet 1 (p = 0.019).  The post-hoc Tukey test did not find a significant difference between the mean weight loss of diet 2 and diet 1 (p = 0.999).  These results should be treated with caution.  The weight loss of diets 2 and 3 are not entirely symmetrical and the variance of weight loss between the three diets is not entirely equal.  In addition, the response variable (weight loss in kg) is only somewhat normal (see exploratory plots above).  Despite this, we did achieve a significant result so diet does have some effect on weight loss.</span>

 
 
 
 How much of the variation is explained?  If you notice when you do a summary from a `lm()` (regression) model, it gives you the $R^2$ values directly (remember $R^2 = \frac{SS_{model}}{SS_{Error}}$).  Does `summary()` of your `aov` model give you that?
 
```{r}
summary(fit.diet)
```

<span style="color: red;">The summary of the aov model does not provide the $R^2$ values directly.</span>

Since I asked the question, the answer is probably no.  Why does it not do this?  Probably for historical reasons, which are a bit of a pain in the backside.  That being said, there are some `tidy` ways to fix this issue.  I'm going to use the `broom` package which allows us to clean up (or tidy if you will) the model objects.  This will take the model object and pull out *all* the ANOVA table stuff and put it into a `tibble`.

```{r}
library(broom)
tidy_diet <- tidy(fit.diet)
tidy_diet
```

Now, since it is all easily accessible, we can calculate the $R^2$ from the new model output.

```{r}
# Estimate the variance explained from the raw sums of squares
r2_Diet <- tidy_diet$sumsq[1] / sum(tidy_diet$sumsq)
r2_Diet 
```



*Gender:* 

The underlying linear model.


$y_{ij} = \mu + \tau_{gender, i} + \epsilon_j$



Independent of the diet, test the null hypothesis $H_O:$ *There is no difference in weight loss between genders* (e.g., $\tau_{gender-0} = \tau_{gender-2} = 0.0 $).  Is there evidence for one gender being significantly different than another?  How much of the variation is explained (another $R^2$ by gender)?

```{r}

fit.gender <- aov(weight.delta ~ gender, data = data)
anova(fit.gender)

gender.sum <- data %>% 
  group_by(gender) %>% 
  summarise(MEAN = mean(weight.delta),
            N = length(weight.delta),
            SD = sd(weight.delta),
            CI= 1.96 * sd(weight.delta) / sqrt(length(weight.delta)))
gender.sum
```

<span style="color: red;">Based on this ANOVA, I fail to reject the null hypothesis (F = 0.0438, DF = 1, p = 0.835) and conclude that the average weight loss of neither gender is significantly different from the other.  The mean weight loss of gender 0 (3.893 +/- 0.752) was lower than the mean weight loss of gender 1 (4.015 +/- 0.863) but as the ANOVA revealed, this difference is not significant.  These results should be treated with caution.  The weight loss of genders 0 and 1 are not entirely symmetrical and the variance of weight loss between the two genders is not equal.  In addition, the response variable (weight loss in kg) is only somewhat normal (see exploratory plots above).</span>


# Do genders respond differently to diets?


$y_{ijk} = \mu + \tau_{Diet,i} + \tau_{gender,j} + \epsilon_k$

So here we do something a bit different.  We want to simultaneously ask the following questions:

- Do diets influence weight loss?
- Do genders influence weight loss?
- Is there an interaction where different genders respond differently to different diets?

In $R$, this is done as:

```{r}
anova(fit.diet)
```

```{r}
# gender model
anova(fit.gender)
```

```{r}
# interaction model
fit.dietgender <- aov(weight.delta ~ Diet*gender, data = data)
anova(fit.dietgender)
```

What is the $R^2$ for this model?

```{r}
# best model variance (diet alone) explained

tidy_dietgender <- tidy(fit.dietgender)
r2_diet <- tidy_dietgender$sumsq[1] / sum(tidy_dietgender$sumsq)
r2_diet

# second best model variance (diet and gender interacting) explained

r2_dietgender <- tidy_dietgender$sumsq[3] / sum(tidy_dietgender$sumsq)
r2_dietgender

# total variance explained by this model
r2_total <- (sum(tidy_dietgender$sumsq) - tidy_dietgender$sumsq[4]) / sum(tidy_dietgender$sumsq)
r2_total

```


# Which Model is Best?

How would you compare the models you generated?  How do you interpret the findings?

*Explain, in words, your findings and interpretation of these findings*

<span style="color: red;">The total variance explained by diet, gender, and the interaction of the two is 20.088%.  Of that total, the results of our ANOVA gave one very significant result showing 12.853% is explained by diet alone.  The ANOVA gave another significant result though it is only barely significant - that 7.200% is explained by the interaction of diet and gender.  My interpretation of these findings is that diet - particularly whatever was in diet #3 - has an impact on weight loss.</span>

