---
title: "Correlation Assignment"
format: html
author: "Candice Weber"
date: today
abstract: "Examining the correlation of various atmospheric data collected at the Rice Rivers Center in January and February 2014." 
---


```{r message = FALSE, warning = FALSE}
library(tidyverse)
library(lubridate)
library(GGally)
source("get_rice_data.R")
data <- get_rice_data()
```

1. Look up the library GGally; it has a function named ggpairs(). Use that function to plot the atmospheric data from the Rice Rivers Center for the samples collected from 10 February to 15 February. 

```{r}
days <- c(10, 11, 12, 13, 14, 15)
data %>% 
  filter(Month == "February") %>% 
  filter(Day %in% days) %>% 
  select (RecordID, Month, Day, AirTempF, RelHumidity, BP_HG) -> atmospheric

ggpairs(atmospheric, columns = 4:6)
```

2. For those atmospheric data, which pair of variables has the strongest correlation? What is the 95% confidence interval on that correlation coefficient?

The pair of variables with the strongest correlation is barometric pressure and air temp.  The 95% confidence interval is -0.750 and -0.669.

```{r}
cor.test(atmospheric$AirTempF, atmospheric$BP_HG)
```


3. Using the first 40 observations in air temperature and barometric pressure from the Rice Center data set, determine if they are individually distributed as normal random variables.  

```{r}
data %>% 
  filter(row_number() <= 40) -> top_40

# Determining if air temp is individually distributed as a normal random variable
shapiro.test(top_40$AirTempF)

# Determining if barometric pressure is individually distributed as a normal random variable
shapiro.test(top_40$BP_HG)

```

The variable of air temperature is normally distributed.  I fail to reject the null hypothesis of the Shapiro-Wilk normality test.  P-value for air temperature is 0.2031 which is greater than 0.05.

The variable of barometric pressure is not normally distributed.  I fail to reject the null hypothesis of the Shapiro-Wilk normality test.  P-value for barometric pressure is 0.0016 which is less than 0.05.




4. Given your findings in the last question, what kind of correlation statistic would be most appropriate for estimating the correlation between this subset of data?

A Spearman's rank correlation test would be appropriate because it does not need both variables to be normally distributed.

```{r}
cor.test(top_40$AirTempF, top_40$BP_HG, 
                            method = "spearman" )
```


5. Look at a qqnorm() plot of the barometric pressure data you used in the previous example. Is there something that "looks" odd with these data? Explain why those data are the way they are.  

```{r}
qqnorm(top_40$BP_HG)
qqline(top_40$BP_HG, col="red")
```

These data look like a staircase because there is a lack of variability in their values.  They cluster around just a few integers.



6. Using a permutation approach, define the distribution of correlation values between the variables in #3, assuming that the NULL hypothesis is true. Plot these as a histogram and include the observed correlation. 

```{r message = FALSE, warning = FALSE}
x <- top_40$AirTempF
y <- top_40$BP_HG
df <- data.frame(Estimate = factor(c("Original",
                                        rep("Permuted", 999))), 
                  rho = c(cor.test(x, y)$estimate,
                            rep(NA, 999)))

for(i in 2:1000) {
  yhat <- sample(y,
                  size = length(y), 
                  replace = FALSE)
  model <- cor.test(x, yhat)
  df$rho[i] <- model$estimate 
}

ggplot(df) + 
  geom_histogram(aes(rho, fill=Estimate))+
  geom_text(aes(x = -0.85, y = 100, 
            label = "Observed correlation: \n-0.5572"), size = 5,
            hjust = 0, color = "black")+
  labs(x = "rho", 
       y = "Frequency", 
       title = "Permutation Testing Air Temperature \nand Barometric Pressure for Significance")+
  theme_minimal()

```

The observed correlation falls just outside of the permuted correlation estimates.  I reject the null hypothesis.  The correlation is not zero.