---
title: "Raster Data Homework"
format: html
author: "Candice Weber"
date: today
abstract: "Loading, manipulating, and displaying geospatial data in a raster format."
---

## 1. Load the raster and point data in and crop to an appropriate size to display the locations of the sampling plots and make a label for each site.

```{r message = FALSE, warning = FALSE}
library(raster)
library(tidyverse)
library(ggrepel)
library(sf)

raster_url <- "https://github.com/DyerlabTeaching/Raster-Data/raw/main/data/alt_22.tif"
beetle_url <- "https://raw.githubusercontent.com/DyerlabTeaching/Raster-Data/main/data/AraptusDispersalBias.csv"

r <- raster(raster_url)
beetles <- read_csv(beetle_url) 
beetles %>%  
  st_as_sf(coords = c("Longitude","Latitude"), crs ="4326") -> beetles_sf

beetles_bbox <- c(-116, -109, 22, 30) 
baja_extent <- extent(beetles_bbox)
alt <- crop(r, baja_extent)
rm(r)
alt %>%
  rasterToPoints() %>%
  as.data.frame() %>% 
  transmute(Longitude=x,
            Latitude=y,
            Elevation=alt_22)  -> alt.df
alt.df %>%
  ggplot()  + 
  geom_raster( aes( x = Longitude, 
                    y = Latitude, 
                    fill = Elevation) ) + 
  geom_sf( aes(), 
           data = beetles_sf, 
           color = "blue",
           alpha = 0.75)+
  geom_text_repel( aes( label = Site,
                        geometry = geometry),
                   data = beetles_sf,
                   stat = "sf_coordinates", 
                   size = 4, 
                   color = "blue") +
  coord_sf()+
  scale_fill_gradientn( colors=terrain.colors(100))+
  theme_minimal()
```

## 2. Use the click() function to crop the raster and filter the sites to include only the sampling locations that are in the Cape regions (e.g., south of site labeled #75).  Plot the raster of elevation and all cape region locales with sex-ratio represented as the size of the plot symbol.

```{r}
# plot( alt )
# click(alt, xy=TRUE, n=4 ) -> points

cape_bbox <- c(-111.4, -108, 22, 24.7)
cape_extent <- extent(cape_bbox)
cape_alt <- crop(alt, cape_extent)

cape_alt %>%
  rasterToPoints() %>%
  as.data.frame() %>% 
  transmute(Longitude=x,
            Latitude=y,
            Elevation=alt_22)  -> cape_alt_df

cape_sites <- c("163", "153", "48", "157", "73", "Aqu")
beetles_sf %>% 
  filter(Site %in% cape_sites) -> cape_beetles

cape_alt_df %>%
  ggplot()  + 
  geom_raster( aes( x = Longitude, 
                    y = Latitude, 
                    fill = Elevation) ) + 
  geom_sf( aes(size = MFRatio), 
           data = cape_beetles, 
           color = "blue",
           alpha = 0.75)+
  geom_text_repel( aes( label = Site,
                        geometry = geometry),
                   data = cape_beetles,
                   stat = "sf_coordinates", 
                   size = 4, 
                   color = "blue") +
  coord_sf()+
  scale_fill_gradientn( colors=terrain.colors(100))+
  theme_bw()


```

## 3. The peninsula of Baja California is divided into the States of Baja California Norte and Baja California Sur. The border between these states is at 28° Latitude. Divide the sample locations into groups based on which state they are located in and plot the average sex ratio of the sample sites partitioned by each site.  

```{r}
beetles %>% 
  mutate(State = case_when(
    Latitude >= 28 ~ "Norte",
    Latitude < 28 ~ "Sur")) -> state_beetles

state_beetles %>% 
  ggplot(aes(x = State,
             y = MFRatio,
             fill = State))+
  geom_boxplot()+
  scale_fill_manual(values = c("lightgreen", "lightblue"))+
  labs(x = "State",
       y = "Sex Ratio",
       title = "Average Sex Ratio by State")+
  guides(fill = "none")+
  theme_minimal()

```

## 4. Is there a relationship between the observed sex ratio and the elevation at that site?  Plot these, and in the text, perform a correlation test (cor.test()) and report the values as if you were writing them up in a results section.

```{r message = FALSE}
elevation_values <- raster::extract(alt, as(beetles_sf, "Spatial"))
beetles_sf$Elevation <- elevation_values

beetles_sf %>% ggplot(aes(x = Elevation,
                 y = MFRatio)) +
  geom_point()+
  geom_smooth(method = "lm", se = FALSE, color = "blue") + 
  labs(x = "Elevation",
       y = "Sex Ratio",
       title = "Correlation of Sex Ratio and Elevation")+
  theme_minimal()


cor.test(x = beetles_sf$Elevation,
         y = beetles_sf$MFRatio)
```
A correlation test of site elevation and site sex ratio resulted in a Pearson's correlation coefficient of -0.1033 (p = 0.5801, df = 29).  There is a slight negative correlation between site elevation and sex ratio.  As the elevation of the site increases, there is a slight decrease in the male-to-female ratio at that site.  Sites with higher elevations were found to have more females to males than sites with lower elevations.  However, the correlation is very weak and the p-value of 0.5801 is much greater than 0.05.  This means there is no significant relationship between these two values.  
